Measuring Stability and Change in Personal Culture Using Panel Data

### 2019-12-06 old_drafts/manuscript7.Rmd
kevin's revisions

---
title: "Measuring Stability and Change in Personal Culture Using Panel Data"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-01.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad models of individual change. One approach theorizes that people actively update their beliefs in the face of new information. The other argues that, following early socialization experiences, dispositions are stable. We formalize these two models, elaborate empirical implications of each, and derive a simple combined model for comparing them using panel data. We test this model on more than 184 attitude and behavior items from the 2006-14 rotating panels of the General Social Survey. Though the pattern of results is complex, it is somewhat more consistent with the settled dispositions model than the active updating model. Most observed change in the GSS appears to be short-term attitude change or measurement error rather than persisting changes. When persistent change occurs, it is somewhat more likely to occur in younger people than older people and more common for public behaviors and beliefs about high-profile issues than private attitudes. We argue that we need both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.
---

```{r setup, include=FALSE}
# Before the first commit, I added "here", changed the load paths with here(), and added the %!in% function

knitr::opts_chunk$set(echo = FALSE)
library(ggdag)
library(ggrepel)
library(here)
library(tidyverse)
'%!in%' <- function(x,y)!('%in%'(x,y))

@@

Attempts to account for opinion changes in society have produced conflicting theories about the nature of opinion formation and the ability of individuals to maintain consistent attitudes. Some models suggest that people lack the cognitive tools to maintain consistent beliefs on social and political issues. As a result, they construct responses on the fly in the interview setting, drawing on ideas from opinion leaders and changing their attitudes as elite discourse changes [@zaller; @converse; @perrinmcfadden]. By contrast, cohort replacement theories posit that people do hold opinions, are unwilling to alter them in the face of societal change, and thus public opinion changes only with generational turnover [@mannheim; @ryder]. Another set of models claim that people hold "a number of real, stable, and sensible opinions about public policy" and change their opinions in response to new information [@pageshapiro: p. xi; @achen1992]. More recently, sociologists have suggested that people attempt to align malleable peripheral beliefs with relatively fixed "core" beliefs using social cues [@boutylinevaisey; @lakoff; @goldbergstein].

Despite their diversity, models of aggregate attitude and behavior change often implicitly invoke one of two broad models of individual change with strong connections to major theories in cultural sociology. The first of these models is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions on attitude formation. This model is rooted in a broadly pragmatist approach to action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. This model has affinities with the Bourdieusian tradition, which de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016). Some models of aggregate opinion change invoke just one of these models, while others suggest that some beliefs follow one model while other beliefs follow another (CITE???). Separate from this distinction is a second dimension of attitude formation that reflects degree to which people can consistently report the same opinion over time. Opinions can be settled but difficult to report, such as when individuals construct responses to the same question over time from a settled but diverse set of social cues [@zaller]. Attitudes can also be in a process of active updating but relatively easy to report, such as when a voter changes her registered partisan affiliation.

In this paper, we make these two models of attitude change explicit, deduce some of their empirical implications, and derive a simple strategy for estimating the prevalence of active updating using panel data. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). By focusing on the pattern of attitude change at the individual level, we can describe the overall pattern of attitude change in the population and clarify when different accounts of aggregate opinion change are more likely to apply. Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary U.S. society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in attitude or practice probably reflects short-term (i.e., non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, settled opinions vary in how consistently individuals report the same answer. Consistent with theories that argue that people lack clear opinions, some survey items appear to elicit inconsistent responses and a few appear to elicit random responses. At the same time, people are consistent on many both high- and low-profile public policy items, suggesting a greater degree of "real" attitudes than these theories suggest. Third, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. On several items it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Fourth, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology), and people are more likely to report these attitudes consistently. This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

These findings are generally somewhat more consistent with the settled dispositions model than with the active updating model. This is, they offer broad support to theories claiming that cultural change tends to come through generational turnover rather than persuasion. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change and that challenge singular existing theories of attitude change. The pattern of results also supports models of attitude change that put ideological identification at the center of a network of political beliefs and suggest that individuals are more likely to make lasting changes in their partisan identification that to most specific policy positions. We argue that there is a place for both models in our theory of cultural evolution but that we need more research on the circumstances under which each is more likely to apply.

@@

## Belief Formation in Cultural Sociology

@@

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions at all. We will explain this in greater detail below. For now, consider the following two simple models:

@@

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from even earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property, and this active updating model generally, is often assumed to underlie the data generation process in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings]. Concerns about panel conditioning, in which a person's participation in a survey influences her subsequent answer choices, either by changing her non-survey behavior or by altering how they respond to the survey without changing behavior, also imply some sort of active updating on the part of the respondent [@warrenhalpernmanners].

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, and this model underlies numerous theories of rational updating in the face of new information [@achen1992], but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

@@

Equation \@ref(eq:FE) represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are thus temporary influences only.

@@

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that these shifts will be erased over time as people return to their baselines.

@@

The models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline over time. In the SDM, changes tend to revert because each person returns to their existing baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_. We consider the implications of these predictions below.

## Consistency in Response
Neither model makes any assumptions about the magnitude of of the variance of $\nu_{it}$, meaning that both models allow for the possibility of either high or low levels of change in responses from wave to wave. This means that we must consider separately the overall _amount_ of change in responses from wave to wave and the _persistence_ of this change. Knowing that there is more wave-to-wave change in one variable than another tells us little about how these two items relate to underlying theories of attitude development. However, knowing that two attitudes have the same rate of wave-to-wave change, but that one follows an active updating pattern with little random fluctuation, while the other follows a settled disposition pattern with a high amount of random fluctuation allows us to make more precise inferences about the applicability of different theories to these items.

@@

These two dimensions of attitude change give us a two-by-two matrix of patterns for attitude change. The first dimension reflects what we call active updating, or the idea that more recent responses are better predictors of current responses that older responses. In other words, active updating means that when changes happen, they are more likely to persist than revert. We note here that a high degree of active updating does not imply a high _rate_ of active updating, simply the predictive power of new responses compared to older responses. People might change attitudes infrequently, but once they do, their new positions are substantially better at predicting future positions than their prior positions. The second dimension reflects what we call stability, or how consistent people are in reporting their beliefs once we account for the amount of active updating.

With these two dimensions outlined, we can begin to categorize predictions of existing theories of attitude development with respect to them. Table ## sketches these two dimensions and outlines where a number of theoretical propositions align in this matrix. We consider these theories in more detail below but note at the outset that our goal here is not to make an explicit prediction for the kind of attitude change we will see for all kinds of attitudes. Instead, we seek to identify sources of tension in these existing models and provide a lens through which to interpret the overall patterns of attitude change across a wide range of types of attitudes.

\begin{center}
\begin{tabular}{ | c | p{4cm} | p{4cm} | } 
\hline
 & Low Active Updating & High Active Updating \\ 
\hline
High Consistency & Attitudes shaped by cohort effects (older individuals); High-profile, settled public policy questions; Core beliefs & High-profile, changing public policy questions; Public attitudes and behaviors; "Rational" beliefs and behaviors \\ 
\hline
Low Consistency &  \\ 
\hline
\end{tabular}
\end{center}

### Age and Attitude Formation

In broad terms, cohort replacement theories of public opinion change posit that individuals hold relatively stable opinions, that few individuals change their attitudes as they age, and that most change in public opinion must come from older individuals dying and being replaced in the population with young individuals who hold different opinions [@mannheim; @ryder; @vaiseylizardo]. Implied in these models is the notion that individuals have a period of "cohortization" in which they form these initial beliefs.

To continue the metaphor above, cohort replacement theories imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition). The concept of a "cohort effect" is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman]. Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative [@vaiseylizardo]. To put it slightly differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

There is abundant evidence that adolescence and early adulthood are periods of heightened attitude change, with this change diminishing rapidly or gradually, depending on the attitude [@glenn; @inglehartbaker; @visserkrosnick]. Most theories agree that middle age (40-60) is a time of reduced attitude change, but some also posit that older individuals might also have increased susceptibility to attitude change [@visserkrosnick; @danigelis].

Many of these studies focus on common measures of political attitudes such as ideological and partisan identification [@krosnickalwin], but Vaisey and Lizardo [-@vaiseylizardo] suggest that this pattern should be the case for a broad range of general attitudes about race and gender. Danigelis, Hardy, and Cutler [-@danigelis] also provide evidence that attitudes about economic gender equality and structural reasons for racial inequality see higher levels of "intracohort" change, a good indicator that some cohort members are changing their minds, in younger cohorts than older cohorts. Theories and evidence of secularization also suggest that most change in religious attitudes and behaviors comes through cohort replacement, rather than individuals changing their beliefs throughout their lives [@voaschaves].

As Table ## suggests, items that reflect a cohort replacement pattern should demonstrate low active updating at the individual level for most adults in the population, but high active updating for young people. Because these approaches generally assume that beliefs are prominent in the minds of respondents, the implication is that these beliefs are reported with high consistency in adulthood. At younger ages, however, we expect more volatility as attitude development occurs. We note, however, that there is some evidence that volatility of settled beliefs increases as people change as well [@krosnickalwin], but we do not expect this to be the general pattern.

### Inconsistent Opinions

Not all cultural change occurs through cohort replacement processes, and there is abundant evidence of large-scale public opinion shifts that occur in a span of time too short to be driven entirely by cohort replacement. This suggests that, at times, even middle-aged individuals update their opinions in persisting ways.

In contrast to cohort replacement theories, a range of theories posit that, outside of a well informed elite, individuals lack consistent opinions and are unable to report them as such in a survey context [@perrinmcfadden; @bourdieu_publicopinion]. These theories tend to focus on questions of public policy, since the act of opinion creation requires much more information than individuals typically have [@zaller; @converse]. For example, when asked whether the government should spend more on health care, forming an opinion requires someone to understand the current landscape of federal spending on health care, issues in the health care system, the potential impact of increased spending on these issues, and the benefits that he might accrue as a result. Most people do not have this information readily stored in their minds.

Because of the cognitive demands of holding consistent opinions on these issues, these theories posit that individuals simply generate responses in the interview context, drawing on resources from their immediate social environment. This explains why survey respondents are susceptible to question-order, priming, and recency effects [@zaller]. However, these theories suggest that when questions of public policy receive significant media attention, individuals obtain more information and can more clearly articulate their positions on these issues. Shifts in public opinion can also come about when elites shift their opinions on issues, providing new signals to the general public.

In general, then, we expect low-profile and complex public policy issues to show a low rate of active updating and a low rate of consistency. By definition, most questions of public policy should fall into this group, since only a few issues can sustain a relatively high level of public attention. The small number of high-profile issues might demonstrate a higher rate of consistency, but we do not expect high rates of active updating for most of these. However, we do spotlight one issue that saw significant change in elite opinion. In the time frame we observe (2006-2014), there was a major shift in elite opinion on the issue of gay rights. Leaders of the Democratic Party (including President Barack Obama) came out in favor of legalizing gay marriage, and a number of Republican Party leaders also began to express support for gay marriage. For this issue, then, we might expect that public opinion shifts notably. This should manifest as a high degree of active updating for the whole population, not just young people.

Outside of a lack of information on public policy issues, there are two other reasons why individuals might demonstrate low consistency and low active updating. First, individuals might be unwilling to report socially undesirable beliefs if they hold them, and as a result obfuscate their true position. Similarly, some issues might not require a lot of information to comprehend, but might be so far from the public consciousness as to be nonsensical to respondents. We expect both of these conditions to produce a similar pattern of low active updating and low consistency.

### Updating Opinions

The third broad set of opinion formation processes posit that rather than elite leadership that determining a belief's malleability, people update their opinions across their lives in response to gathering new information and interacting with changing social environments.

Some of these approaches, common in political science, suggest that people consistently incorporate new information to refine their opinions over time [@achen1992; @bartelsjackman]. These models do not tend to make distinctions about which kinds of attitudes shift over time and which do not. However, for these models to be true, we would have to observe a high rate of active updating and a high rate of consistency.

Other theories suggest that only some beliefs are open to updating during adulthood. A number of theories of attitude development posit that people hold “core” beliefs that they use to adjudicate more peripheral attitudes, such as views on specific policy questions they face in an interview context [@boutylinevaisey; @goldbergstein]. These approaches include Lakoff’s [-@lakoff] model of moral politics, where views of the family serve as a template onto which to transpose questions of public policy, and Converse’s [-@converse] notion of social constraint, where political identity is used to form more specific policy stances as a result of information streams. In their analysis of cross-sectional data, Boutyline and Vaisey [-@boutylinevaisey] find little support for Lakoff's model and significant support for ideological identification as this "core" belief in the political space. If these individuals have the ability to regularly assess new information against these core beliefs, they should also be able to recognize their position in a multiple-choice survey questionnaire and report them with limited error [@vaisey09]. In other words, we expect that core beliefs such as ideological identification should have high consistency and low active updating, and conversely that items that show these traits to have the potential to be core beliefs.

In a similar vein of research focused on political polarization, Baldassari and Gelman [@baldassarigelman] find little increase in the correlation between specific policy positions over time, suggesting limited active updating of these items. What they do find is an increased alignment between partisan identification and policy positions. They interpret this finding as indicating that "voters are splitting along party lines according to the issues that are most salient to them, while they do not bother to adjust their (weak) preferences on the remaining issues" [@baldassarigelman: pp. 442]. This suggests that among items tapping issues of politics, partisan affiliation should demonstrate the highest level of active updating, and that it should be reported with a high degree of consistency.

### Public vs. Private Culture

A final dimension that likely influences the consistency of attitude and behavior reports is the publicness of the item, or whether it is in some way externalized in public symbols, discourses, and institutions. The act of opinion construction often taps what Lizardo calls "personal declarative culture," the "explicit, symbolically mediated culture" such as language that people use to reason, evaluate, judge, and categorize objects [-@lizardo17: pp. 91]. This kind of knowledge is contradictory and flexible, meaning a person can reach and justify a range of different opinions in response to the same prompt, depending on the circumstances, which might lead to a low degree of consistency and active updating. However, when public culture provides a strong "scaffold" for individuals through clear signals of how their identities and social locations should influence how they should respond to a survey question, it becomes easier for them to maintain consistency [@lizardostrand].

We have touched on this idea already with our discussion of the idea that high-profile elite opinion change might drive opinion change in the broader public, or by noting that high-profile issues might be easier to report consistently because individuals receive signals that support that position. Along the same line, items that tap some public dimension of individual behavior and attitudes are likely to show higher levels of active updating because social mechanisms, such as joining a group, can help maintain changes in ways that intrapsychic forces cannot. Any item where the burden of maintaining consistency is externalized and not only "in a person's head" might demonstrate both higher rates of active updating and greater consistency.

### Toward Theoretical Synthesis

The clearest contrast in existing theories is between models that assume that beliefs are relatively settled during adulthood (cohort replacement theories), which would produce a pattern of low active updating, and those that posit that individuals update beliefs throughout their lives (most other models), which would produce a pattern of high active updating. This disagreement thus forms the primary structure of our analysis. A lack of evidence in favor of active updating for most individuals in the population, coupled with evidence of active updating in younger members of the population and a high degree of consistency, would generally suggest that individuals form opinions early in life and maintain them throughout adulthood, consistent with cohort replacement and habitus theories of attitude development. We should note that it is possible that individuals have finished their developmental period prior to reaching adulthood, in which case we might not observe age differences in active updating. However we should still observe high consistency and low active updating.

If what we find is a diverse array of patterns, with items displaying a range of active updating and consistency, then the above theories provide a window through which to interpret these distinctions. Our goal here is neither to declare victory for one of the theoretical perspectives nor to simply say that all "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types of processes are at work. We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

@@

Third, _are there systematic differences in item content between questions that fall in different parts of our matrix?_ The preceding sections have made some specific predictions based on existing literature, but we cannot enumerate predictions for all sorts of beliefs. As we note, our approach is to use these theories as a lens through which to interpret the overall pattern of results.

@@

To investigate these ideas empirically, we examine 184 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below. While a broader range of years would be preferable, the panel component of the GSS began in 2006 and was discontinued in 2014, so these years represent the full range of what we could analyze using the GSS.

@@

Our main goal is to estimate separately the amount of active updating and the amount of non-persisting change in responses over time. We first consider measuring the amount of active updating. Our two different models make different predictions in the three-wave panel context. The AUM makes the following prediction for wave 3:

@@

```{r}
load(here("data", "results_df.Rdata"))
load(here("data", "label_vars.Rdata"))
load(here("data", "aum_vars.Rdata"))

>>>>

---
title: "Measuring Stability and Change in Personal Culture Using Panel Data"
author:
- Kevin Kiley
- Stephen Vaisey
date: "Duke University"
output:
  bookdown::word_document2: 
    fig_caption: yes
    reference_docx: word-styles-reference-01.docx
  pdf_document:
    number_sections: yes
header-includes: \usepackage{longtable}
nocite: |
  @swidler01
bibliography: panel_change_bib.bib
abstract: Models of population-wide cultural change tend to invoke one of two broad
  models of individual change. One approach theorizes that people actively update
  their beliefs in the face of new information. The other argues that, following early
  socialization experiences, individuals' dispositions are stable. We formalize these
  two models, elaborate empirical implications of each, and derive a simple model
  for comparing their prevalence using panel data. We test this model on more than
  184 attitude and behavior items from the 2006-14 rotating panels of the General
  Social Survey. The pattern of results is more consistent with the settled dispositions
  model than the active updating model, and most observed change in the GSS appears
  to be short-term attitude change or measurement error rather than persisting changes
  in belief. When persistent change occurs, it is somewhat more likely to occur in
  younger people than older people and more common for public behaviors and beliefs about 
  high-profile issues than private attitudes. We argue that there is a place
  for both models in our theory of cultural evolution but that we need more evidence
  on the circumstances under which each is more likely to apply.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggdag)
library(ggrepel)
```

@@

Sociological analyses of cultural change tend to focus on population-wide changes in attitudes, beliefs, or practices that, while rarely made explicit, invoke one of two broad models of individual cultural change. The first is an _active updating_ model that emphasizes the role of changing discourses, environments, and interactions. In this model, people continuously revise their views and practices based on the people, discourse, and events around them. The second is a _settled dispositions_ model, which emphasizes the continuing influence of durable dispositions acquired early in life. In this model, adults' views and practices are relatively robust to changes in the social environment.

Both accounts have affinities with traditions in sociological theory. The active updating model is most compatible with a broadly pragmatist approach to social action, which claims that contemporary social environments and problems provoke people to adapt their views and make new meaning (e.g., Joas 1996, Gross 2009, Swidler 2001). The settled dispositions model has affinities with the Bourdieusian tradition that de-emphasizes (though does not ignore) the current environment in favor of the "past conditions of production" (e.g., Bourdieu 1990, Vaisey and Lizardo 2016).

In this paper, we make these two core models explicit, deduce some of their empirical implications, and derive a simple strategy for comparing their prevalence using panel data. We then apply this method to 184 items from the 2006-2014 General Social Surveys (GSS). Because of data limitations, we cannot speak to all types of cultural objects (e.g., music styles, baby names). We do, however, investigate a wide variety of opinions, including views on politics, free speech, race, and gender roles, and practices including socializing at at bars or attending church, that are important in contemporary society.

Our analysis yields several results. First, we find that the majority of what appears to be individual-level change in belief or practice probably reflects short-term (non-persisting) change or measurement error rather than actual persistent change. Simply put, there is little evidence that U.S. adults are changing their beliefs or practices in lasting ways over this period of study. Second, the persistent change that we _do_ see in the data is somewhat more concentrated among younger respondents. That is, on several items at least, it appears that younger adults are still in the process of acquiring dispositions and habits they will take into later life. Third, we find that changes in social behavior (e.g., church attendance, political party membership, socializing) are more likely to be persistent than changes in private attitudes (e.g., political ideology). This suggests that interactional and institutional mechanisms may provide stronger support for lasting change than pressures for intrapsychic consistency.

These findings are generally more consistent with the settled dispositions model than with the active updating model. However, there is a pattern of exceptions and caveats that can help us understand how institutions and events shape the process of cultural change. We argue that there is a place for both models in our theory of cultural evolution but that we need more evidence on the circumstances under which each is more likely to apply.

In addition to their theoretical significance, our findings also have practical methodological implications. We argue that short-duration panel studies of adults will be of limited utility for studies of cultural change unless they have a clear focus on the sorts of things that are likely to change in persistent ways over the study period. Otherwise, researchers will waste resources measuring "change" that turns out to be noise. In general terms, then, we suggest that social scientists would do well to conduct more panel studies of children and adolescents while using cross-sectional designs to study adults.

@@

@@

How does the respondent formulate a response to that question each time, year after year? To make things as explicit as possible, we can write the data-generating process formally. Although this presentation may make it seem like we are assuming rationality or conscious deliberation, this way of writing the models makes no particular cognitive assumptions at all. We will explain this in greater detail below. For now, consider the following two models:

@@

Equation \@ref(eq:AR) represents the active updating model. The respondent forms her answer by starting with what she said last time ($y_{it-1}$) and then incorporating any new considerations ($\nu_{it}$). There is no need to remember responses from even earlier time points (e.g., $y_{it-2}$) because this information gets folded into the updated response each time. Formally, Equation \@ref(eq:AR) is a Markov model, where future states depend entirely on the current state. This formal property, and this active updating model generally, is often assumed to underlie data generation in studies of change in and reliability of repeated survey measures [@alwin; @krosnickalwin; @houthastings].

More informally, this AUM posits an agent that is updating her views in the face of social experience. There are formal Bayesian ways of modeling updating, but we need not rely on any assumptions of rationality, optimality, or conscious thought for this basic process to apply. Following Gross's [-@gross09, pp. 367] pragmatist account, we could instead regard this model as consistent with an "active and creative relation to the world" that "lead[s] actors to see themselves in new ways, to value different kinds of goods, and to become attached to problem solutions that they could not have imagined previously."

@@

Equation \@ref(eq:FE) represents the settled dispositions model. Here each respondent begins the study period with a set tendency to respond to the question in a particular way ($U_i$). Temporary considerations, like current events, can play a role in what response she gives at each time (part of the $\nu_{it}$), but these considerations have no lasting impact beyond time $t$. As the right panel of Figure 1 makes clear, there is no mechanism by which a particular consideration can "propagate up" into the settled disposition and change the baseline for future responses. Such considerations are temporary influences only.

@@

Like the AUM, the SDM does not require $\nu_{it}$ to be 0 in expectation for the population at time $t$, but it does assume that $\nu_{it}$ has an expected value of 0 within an individual over time. That is, current considerations can move a person from their baseline temporarily, but there is a tendency to "bounce back" to that baseline over time. The SDM thus allows for population-wide shifts in beliefs, practices, or identities at a particular time (temporary period effects), but it assumes that within individuals these shifts will be erased over time as individuals return to their baselines.

@@

The models thus differ fundamentally on their emphasis on the character of personal change. In the AUM, changes tend to persist because they propagate into future responses, shifting the baseline. In the SDM, changes tend to revert because each person returns to their existing baseline. We can therefore summarize the models' implications for individual change as follows: the AUM predicts _persisting change_ whereas the SDM predicts _non-persisting change_. We consider the implications of these predictions below.

@@

Although our presentation above highlights differences between these models, they are not necessarily contradictory. There are at least two ways to understand how both models could apply without contradiction.

### Age and Cohort Formation

The easiest way to reconcile the AUM and SDM is to posit that they operate sequentially over a person's life. A range of sociological theories provide conflicting hypotheses and evidence for how susceptible individuals are to attitude and behavior change at different life stages. Most of these hypotheses suggest that early adulthood is a time of heightened susceptibility attitude change. Two hypotheses -- the "increasing persistence" hypothesis [@glenn; @inglehartbaker] and the "impressionable years" hypothesis [@visserkrosnick] -- suggest that susceptibility to attitude change is highest early in life and decreases (gradually or quickly, respectively) as individuals age. The "life stages" hypothesis [@visserkrosnick; @danigelis] posits that both early (<40) and late (60+) life are times of more attitude change but that mid-life is characterized by limited attitude change.

To continue the metaphor above, one can imagine travelers wandering around for a while (active updating) before setting up their permanent home base (settled disposition). This is, in fact, the logic that underlies the notion of a "cohort effect." The idea of a cohort effect is that the contemporary environment influences the baseline response tendency when a person is young and then stabilizes and stays "baked in" for the remainder of one's life [@elder; @bartelsjackman]. Any notion of a settled disposition (including the Bourdieusian one) requires an earlier "formative period" where something like the active updating model was operative [@vaiseylizardo]. To put it slightly differently, cohort formation requires an early period of persistent change followed by a later period characterized by either absolute stability (where the variance of $\nu_{it}$ is low) or by temporary, non-persisting changes that disappear as people revert to baseline.

This discussion of cohorts brings us close to the vast literature on age-period-cohort decomposition models, which is focused on statistical techniques to partition variance among sources of social change in repeated cross-sectional data [@yangland]. Although this work is important, it is focused on modeling changes in population aggregates whereas we are concerned with modeling cultural change from the point of view of the same actor moving through time.

The two are related and have implications for each other, of course. Again, the idea of a cohort requires that some people (i.e., younger people) are influenced by the environment to make persistent change in ways that others (i.e., older people) are not. This implies a transition, perhaps a gradual one, from active updating to a settled disposition. A further connection between the two is that the world implied by the SDM can only change in the aggregate due to cohort replacement. That is, if persisting change is extremely rare among adults, the only way for persisting change to occur in the population is for older cohorts to die and be replaced with younger cohorts with (perhaps) different baselines. For our purposes, the most important idea here is that "cohortization" implies that the AUM might describe the data-generating process better among younger respondents whereas the SDM might describe it better among older respondents.^[We do not have space for a full discussion here, but we regard these approaches as complementary. Repeated cross-section approaches must make strong assumptions about individual processes but are able to use many years of data. A panel approach is better able to capture micro-level processes but at the cost of using only a few years of data; this requires making strong assumptions about the generalizability of the process across historical time.]

### Different Items, Different Processes

Another way to reconcile the AUM and the SDM is to consider that they apply to different sorts of questions a researcher might ask a respondent, or that some attitudes and behaviors are stable while others are constantly updating. We can illustrate this idea with (non-cultural) examples we understand well and then consider how less obvious cultural processes might be similar or different.

First, if we repeatedly ask a sample of women "how many children have you ever given birth to?", we know the data is generated by a process that looks like Equation (1). Leaving measurement error aside, the response is a simple combination of how many children the respondent had last time we asked ($y_{it-1}$) and whether or not she has had any since ($\nu_{it}$).

Second, if we repeatedly ask a sample of adults "how many full siblings did you have at age 16?", we know the true answer cannot change and therefore Equation (2) is the best model. This is because $U_i$ refers to a fixed quantity and cannot change over time. Responses could still vary due to measurement error (e.g., misunderstanding the question) meaning that the variance of $\nu_{it}$ need not be 0. But the underlying reality is, by definition, a settled matter.

Finally, there are responses that have no stable component but are entirely functions of local circumstances. If we repeatedly ask a sample of people in the same city "did it rain yesterday?", their responses would not be described better by Equation (1) or Equation (2). In this case, everything that matters is in the contemporary circumstances ($\nu_{it}$) and thus the terms that distinguish Equation (1) from Equation (2) ($y_{it-1}$ or $U_i$) are ignorable. These types of items are not important for adjudicating our research questions, but we still need to be aware of this possibility.

What we want to determine, then, is how much a particular item resembles these different processes. Consider repeated requests for (dis)agreement with the prompt "a working mother can establish just as warm and secure a relationship with her children as a mother who does not work." Does the aggregate pattern of responses look more like an active updating process where a person might be changing her mind? Or does it look more like asking a person to report repeatedly on a settled matter?

When it comes to beliefs and behaviors, which items will demonstrate persistence likely depends on a number of factors. Vaisey and Lizardo [-@vaiseylizardo] found evidence that beliefs about gender, race, and other "core" opinions are more driven by cohort effects and are therefore less likely to be actively updated by adults. Using repeated cross-sectional data, Danigelis, Hardy, and Cutler [-@danigelis] find that "intracohort" change (a good indicator that some cohort members are changing their minds) is larger for young individuals (18 to 39 years old) than for older individuals (over 60 years old) for attitudes about economic gender equality and structural reasons for racial inequality. For attitudes pertaining to civil liberties, political gender equality, and privacy, cohorts tended to show similar rates of intracohort change at both young and old ages during the time frame they studied.

Previous studies have also found that political attitudes are more volatile during early and late adulthood but stabilize during middle age [@krosnickalwin]. While volatility of responses increased as individuals aged, this is mostly due to increases in measurement error, rather than persisting change. This generally suggests that political views will show decreasing evidence of persistent change as age increases.

"Public culture," such as participating in social activities, may show similar rates of persistence over time because social mechanisms (like joining a group) can help maintain changes in ways that intrapsychic forces cannot. Finally, items that capture issues that are highly salient in the media during the study period may be more likely to show active updating because of heightened exposure. Sears and Valentino [-@searsvalentino] find that during political campaigns adolescents tended to adopt new beliefs about issues that were salient during the campaign but not those issues peripheral to it. Given these different possibilities, we do not begin this study with clear predictions on all issues, but general expectations that items pertaining to private cultural beliefs will show less persistent change and that items related to public cultural expressions will show more persistent change. Because of the large, high-profile discussion on gay marriage and gay rights during this period, however, we do expect to see active updating on related items.

## Toward Theoretical Synthesis

The active updating model and the settled dispositions model are different and they do have affinities with different theoretical traditions in sociology. Indeed, the difference in emphasis between environmental cues, creative meaning-making, and public discourse on the one hand and settled dispositions, "habitus," and cohort replacement on the other has been a basic theoretical tension in cultural sociology for more than a decade [@vaisey09; @lizardo17; @vaiseylizardo].

One model may, in fact, be a better starting place than the other. Previous work using cross-sectional data suggests that settled dispositions are hugely important [@vaiseylizardo]. Despite these findings, the evidence is still mixed [@danigelis]. Therefore our goal here is neither to declare victory for one of the theoretical perspective nor to simply say that both "matter." Rather, our objective is to improve sociological models of cultural evolution by more precisely specifying when and where different types processes are at work. We believe achieving a better understanding of these processes will be relevant for many subfields of sociology as well as for other social science fields that study changes in beliefs and behaviors.

@@

Third, _are there systematic differences in item content between questions that are better described by each model?_ Items differ from each other in many ways and it would be challenging to come up with clear predictions for all possible types of items. In general, however, we expect (following Vaisey and Lizardo) that core beliefs about gender, family, race, and trust should be more consistent with settled dispositions because their formation should be already be complete for adults. Items that require public commitments such as changing political parties, socializing, or religious attendance and affiliation should be open to persistent change because there are social mechanisms (not just cognitive ones) to _maintain_ the change.

@@

To investigate these ideas empirically, we examine 184 survey items from the 2006-2014 General Social Survey panels in search of evidence in favor of an active updating model. This period of the GSS contains three different three-wave panels, each of which surveys a sample of adults three times over a four-year period (e.g., 2008-2010-2012). Three waves of data is the minimum amount needed to compare the predictions of the active updating model and the settled dispositions model. We discuss item selection below.

@@

Our two different models make different predictions in the three-wave panel context. The AUM makes the following prediction for wave 3:

@@

```{r}
load("~/panel_change/data/results_df.Rdata")
load("~/panel_change/data/label_vars.Rdata")
load("~/panel_change/data/aum_vars.Rdata")
```
